Density-dependent selection with strictly relative contests

### 2017-11-15 manuscript.tex
l

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large. The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In this case the $f_1$ and $f_2$ nullclines are not even defined. Moreover, at face value MacArthur's, conclusion that ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} seems to be incompatible with constant $N$. In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in $N$. More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density. This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, using a Lotka-Volterra form for Eq. \eqref{eq:macgeneral}. Assuming the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region. In this example, the outcome depends on initial condition. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

@@

\subsection*{K-selection versus relative fitness}

>>>>

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large. The outcome of selection is then determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In this case the $f_1$ and $f_2$ nullclines are not even defined. Moreover, at face value MacArthur's conclusion that ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} seems to be fundamentally incompatible with constant $N$. In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in equilibrium population size. More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density. This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model in Eq. \eqref{eq:macgeneral}. The type that  (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

\subsection*{$K$-selection, $c$-selection and relative fitness}

Here we compare $K$-selection and MacArthur's broader claims about selection in crowded populations with our density-dependent lottery model. To do so we first need to clarify MacArthur's particular use of ``$K$''.

@@
